Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2023-12-08 manuscript/manuscript.Rmd
Finished implementing Anitra edits to results and intro

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment.[@Karl1999; @Karl2017; @Karl2021] A major source of this variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016]. Prior research has established the importance of these mesoscale features in introducing nutrients into the euphotic zone and altering the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies. @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation, especially at the surface. [Anitra: so the surface is driving changes at the DCM?]

@@

*Figure 1: Plots of the metabolome across the eddy dipole, broken down by depth. The top row of plots are non-metric multidimensional scaling (NMDS) plots, in which points correspond to individual samples and have been colored by their corrected sea level height anomaly. NMDS stress values (s) have been reported in the bottom left corner, while PERMANOVA R$^2$ and p-values are reported in the top left. SLA trends are visible in the DCM and deep samples, with dark blue circles consistently discriminating from the dark red circles along the first multidimensional axis. The bottom row of plots show the direction and magnitude of this effect by plotting the grand mean of the normalized metabolite peak areas in the stations taken across the eddy dipole.*

@@

```{r figure stacked barplot MT eddy transect w PC and SLA}
avg_targ_concs <- targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  group_by(compound_name) %>%
  mutate(mean_cmpd_nm=mean(nM)) %>%
  ungroup()
base_targ_gp <- avg_targ_concs %>%
  group_by(compound_name, untripl) %>%
  summarise(nM=mean(nM), mean_cmpd_nm=unique(mean_cmpd_nm)) %>%
  ungroup() %>%
  arrange(desc(mean_cmpd_nm)) %>%
  mutate(compound_name=ifelse(compound_name=="Dimethylsulfoniopropionate", "DMSP", compound_name)) %>%
  mutate(compound_name=ifelse(
    compound_name%in%head(unique(compound_name), 11), 
    compound_name, "Other")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  left_join(filled_file_metadata %>% distinct(untripl, depth, station)) %>%
  group_by(depth, station, compound_name) %>%
  summarise(nM=sum(nM), .groups = "keep") %>%
  summarise(mean_nM=mean(nM)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=station, y=mean_nM, fill=compound_name), width = 0.9) +
  scale_fill_manual(values = c(hcl.colors(12)[c(1,3,5,7,9,11,2,4,6,8,10)], "grey50")) +
  scale_y_continuous(expand = c(0.005, 0.1)) +
  labs(x="Station", y="Mean nM metabolite concentration", fill=NULL) +
  facet_grid(~depth, scales="free", space = "free") +
  theme_bw() +
  theme(legend.position = "bottom")

sla_loess_points <- filled_file_metadata %>%
  filter(samp_type=="Smp") %>%
  filter(ms_run=="MT") %>%
  distinct(station, sla_corr) %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  group_by(station) %>%
  summarise(sla_corr=mean(sla_corr)) %>%
  mutate(station=as.numeric(station)) %>%
  loess(formula = sla_corr~station) %>%
  predict(., newdata=data.frame(station=seq(1, 11, 0.01))) %>%
  data.frame(station=seq(1, 11, 0.01), sla_corr=.)
arrow_df <- data.frame(
  cyc_type=c("Anticyclone\ncenter", "Cyclone\ncenter"),
  x=c("6", "12"), xend=c("6", "12"),
  y=c(5, 4), yend=c(7.5, 6.5),
  color=c(scales::muted("red"), scales::muted("blue")),
  depth=factor("Deep", levels=c("Surface", "DCM", "Deep"), 
               labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)")),
  lab_x=c("5", "11")
)
targ_gp_w_sla <- base_targ_gp +
  ggnewscale::new_scale_fill() +
  geom_tile(aes(x=station, y=-0.5, fill=sla_corr),
            color=NA, width=0.11, height=0.5, data = sla_loess_points) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), guide = "none"
  ) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend, color=color), data=arrow_df,
               arrow=arrow(ends = "first", length = unit(0.1, "npc"), type = "closed"),
               linewidth=0, linejoin = "bevel") +
  geom_segment(aes(x=x, y=y+0.5, xend=xend, yend=yend, color=color), data=arrow_df, linewidth=2) +
  geom_text(aes(x=lab_x, y=yend+0.5, color=color, label=cyc_type), 
            data=arrow_df, vjust=0, hjust=0, lineheight=0.8) +
  scale_color_identity()

PC_data <- filled_file_metadata %>%
  filter(samp_type=="Smp") %>%
  filter(ms_run=="MT") %>%
  distinct(station, PC_um, depth) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  mutate(station=factor(station, levels=as.character(4:14)))
targ_gp_w_sla_PC <- targ_gp_w_sla +
  geom_point(aes(x=station, y=PC_um*5), data=PC_data, pch=21, fill="white", color="black") +
  scale_y_continuous(sec.axis = sec_axis(~./5, name = "Particulate carbon (o, µM)"),
                     expand = c(0.005, 0.1))

ggsave("targ_gp_w_sla_PC.png", plot = targ_gp_w_sla_PC, device = "png", 
       path = "figures", width = 6.5, height = 4.5, units = "in", dpi = 300)
```

![](figures/targ_gp_w_sla_PC.png)

*Figure 2: Concentrations of known compounds across the eddy dipole, broken down by depth. Stacked bars correspond to known compounds, with the top 11 shown and all other metabolites summed in grey (n=[TBD]). DMSP = dimethylsulfoniopropionate. Sea level anomaly (SLA) is shown by the red-blue color scale underneath the barplot with red corresponding to a positive SLA (anticyclone) and blue corresponding to negative SLA (cyclone). The eddy centers (stations 6 and 12) have been labeled with arrows of the same color scheme in the final panel. Particulate carbon concentrations in µmol/L are shown as white dots and use the y-axis scale on the right side (200x the metabolite scale) [Anitra: This implies that the metabolites are 1/200 of the POC. But you have plotted compound concentration not carbon for metabolites. Assuming each compound has an average of 4 carbon atoms, you would have to multiply by 4 so that is 1/50 (2%) or so of POC. I think you need to make this clear somehow. Or you could make an additional plot with C for each compound. I see supp fig 2 has the %POC so maybe that is enough but maybe call out that figure here.][Will: I agree, let's brainstorm].*

```{r avg targ conc values and numbers for text, eval=FALSE}
avg_targ_concs %>%
  group_by(filename) %>%
  summarise(nM=sum(nM)) %>%
  left_join(filled_file_metadata %>% distinct(filename, untripl, PC_um, depth)) %>%
  group_by(untripl, PC_um, depth) %>%
  summarise(nM=mean(nM), .groups = "drop") %>%
  group_by(depth) %>%
  summarise(PC_vs_nM=broom::tidy(cor.test(PC_um, nM)))
  # summarise(PC_vs_nM=broom::tidy(cor.test(PC_um, nM, method = "spearman")))
  # ggplot(aes(x=PC_um, y=nM, color=depth)) + geom_point() + geom_smooth(method="lm")
avg_targ_concs %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  group_by(depth, compound_name) %>%
  summarise(nM=mean(nM)) %>%
  arrange(compound_name) %>%
  pivot_wider(names_from = depth, values_from = nM) %>%
  write_csv("supplement/avg_targ_conc_by_depth.csv")
avg_targ_concs %>%
  mutate(depth=factor(depth, levels=rev(c("Surface", "DCM", "Deep")))) %>%
  group_by(compound_name, depth) %>%
  summarise(nM=mean(nM)) %>%
  mutate(nM=log10(nM)) %>%
  ggplot(aes(x=nM, y=depth, group=compound_name)) +
  geom_path()
avg_targ_concs %>%
  mutate(depth=factor(depth, levels=rev(c("Surface", "DCM", "Deep")))) %>%
  group_by(compound_name, depth) %>%
  summarise(nM=mean(nM)) %>%
  mutate(nM=rank(nM)) %>%
  ungroup() %>%
  pivot_wider(names_from = depth, values_from = nM) %>%
  mutate(depth_class=paste(Surface, DCM, Deep)) %>%
  count(depth_class)
```

We found that the overall concentration of known molecules was highest at the DCM of the cyclone, in line with our expectations and prior results. The most abundant quantified molecule overall was trimethylamine N-oxide (TMAO), with an average concentration of 1.38 nM. At the surface hydroxyisoleucine was the most abundant metabolite at 1.53 nM due to its monotonic decrease with depth as opposed to a majority of the osmolytes (37/53) which reached maximum concentrations at the DCM instead (Figure 2).

[Anitra: comments on the paragraph above: Maybe instead of this somewhat convoluted description you could just break it down by depth and say which compound was most abundant at each depth? Or which compound was most abundant in the cyclone and anticyclone. And then in a separate sentence just simply state that most compounds reached a ma at the DCM, but hydroxyisoleucine was an exception. Also, are you implying that hydroxyisoleucine is an osmolyte? Is that true? Also, what do you mean by "average"?]

Bulk particulate organic carbon (POC) measurements correlated well with overall [Anitra: what do you mean by overall here?] metabolite concentrations ($r= .554$, p-value < 0.001) but this was largely driven by both parameters having higher values in the surface than at depth. When analyzed at each depth individually, the DCM ($r_{DCM}= .790$, p-value = 0.003) and 175 meter samples ($r_{Deep}= .554$, p-value = 0.077) had weaker but still good correlations, but no such association was detected at the surface ($r_{Surface}= -.125$, p-value = 0.714).

```{r PC vs metab abund anovas and perm test, eval=FALSE}
pc_comp_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity)) %>%
  mutate(n_C=as.numeric(str_extract(formula, "(?<=C)\\d+"))) %>%
  mutate(nM_C=n_C*nM) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  filter(samp_type=="Smp") %>%
  group_by(filename) %>%
  summarize(total_nM=sum(nM_C), PC_um=unique(PC_um)) %>%
  mutate(frac_in_metab=total_nM/1000/PC_um) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))

# Significance testing between depths and SLA classes for % PC in metab
pc_comp_df %>% aov(formula=frac_in_metab~depth) %>% TukeyHSD()
pc_comp_df %>% aov(formula=frac_in_metab~sla_class) %>% TukeyHSD()

# Significance testing for differences in % in metabs as a function of SLA by depth
pc_comp_df %>%
  nest(data=-depth) %>%
  mutate(aov_out=map(data, ~broom::tidy(aov(.x$frac_in_metab~.x$sla_class)))) %>%
  mutate(kw_out=map(data, ~broom::tidy(kruskal.test(.x$frac_in_metab~.x$sla_class)))) %>%
  pivot_longer(c(aov_out, kw_out)) %>%
  unnest(value) %>%
  drop_na(p.value) %>%
  select(depth, name, p.value) %>%
  group_by(name) %>%
  mutate(p.value=p.adjust(p.value, method="fdr")) %>%
  ungroup() %>%
  arrange(name, depth)
pc_comp_df %>%
  ggplot(aes(x=sla_class, y=frac_in_metab)) +
  geom_boxplot() +
  facet_wrap(~depth, nrow=1)

# Permutation testing for effect size test on Deep SLA differences
set.seed(123)
map_dbl(seq(1, 2, 0.1), function(n_mult){
  map_dbl(1:100, function(h){
    pc_comp_df %>%
      filter(depth=="Deep") %>%
      group_by(sla_class) %>%
      summarize(mean=mean(frac_in_metab), sd=sd(frac_in_metab), n=n()) %>%
      mutate(sim_vals=map(1:3, function(i){rnorm(.$n[i]*n_mult, .$mean[i], .$sd[i])})) %>%
      unnest_longer(sim_vals) %>%
      aov(formula=sim_vals~sla_class) %>%
      broom::tidy() %>%
      filter(term=="sla_class") %>%
      pull(p.value)
  }) %>% mean()
}, .progress = TRUE)
```

This correlation between PC [Anitra: Is there ATP data? Might be interesting to compare. Will: instinct says no, I think that ATP was really only collected during the diel phase and from a few spots during lagrangian. TODO: Double check the dashboard] and overall metabolite abundance implied that the fraction of PC represented by known metabolites was likely to be fixed across the eddy transect and this expectation was validated by followup analyses. [Anitra: I do not see this at all. I see spikes in metabolite concentration that are not matched with spikes in PC. Mostly at the eddy centers. Do you not think this is suggestive of a change in the contribution of biomass to PC in the eddy centers? I would assume if you had biomass instead of PC you would be able to see biomass increase at the centers where metabolites increase. TODO: Check the same values using beam attenuation (interpolate if necessary)] Known metabolites contributed between approximately 1.5% and 5% of the carbon to the total PC load, but this variance was not clearly due to either depth or SLA (Supplemental figure 2). The only difference we observed was a slightly reduced metabolite contribution in the 175 meter samples (0.6 percentage points lower at 175 meters, p-value=0.002) while no difference was detected between the surface and DCM samples or between the cyclone, anticyclone, and outside the eddy samples (all p-values > 0.1). A higher proportion of PC in living biomass would translate to a higher proportion of metabolites in POC. This lack of difference may have been due to the small sample sizes (6 anticyclonic stations, 3 cyclonic, and 2 neither, each sampled in triplicate). A followup permutational sensitivity analysis indicated that a difference in means of 0.4 percentage points would have been obtained at the α < 0.05 level for SLA in the 175 meter samples if ~1.5x as many samples had been collected with mean and variances remaining similar.

[TODO: switch to POC instead of PC after confirming it's only the organics]

  - Discussion: My % PC in metabs lower than Angie's? Because her PC values were way higher but metab concentrations were about the same?

```{r comparison to Angies PC values way higher than mine, eval=FALSE}
angie_pcdata <- "../metadata/raw_data/Karl_HL2_WC_PCPN_Final_Data.xlsx" %>%
  openxlsx::read.xlsx() %>%
  slice(-1) %>%
  select(depth=`Depth.(m)`, PC_um=`C.value.(µmol/L)`) %>%
  filter(depth<30) %>%
  mutate(cruise="HOE-Legacy")

my_pcdata <- filled_file_metadata %>%
  filter(depth=="Surface") %>%
  distinct(depth, PC_um) %>%
  mutate(cruise="MESOSCOPE")

rbind(angie_pcdata, my_pcdata) %>%
  ggplot() +
  geom_boxplot(aes(x=cruise, y=PC_um)) +
  scale_y_continuous(limits = c(0, NA))
```

@@

  - Discussion: I still really like the idea of modeling things according to the slope of their change across the eddy dipole. Light-sensitive or pressure-sensitive (e.g. sunscreen molecules, piezolytes) metabolites wouldn't change at all, DCM-specific metabolites (e.g. chlorophyll) would change by 20m or so as the DCM does, nutrient-sensitive metabolites (guanine? glutamate/glutamine?) would track with the isonutrials, and the salinity or temperature sensitive metabolites (aka pure osmolytes) would track with isopycnals. Do I have enough data to show that here? I forget what my early attempts to quantify these gradients looked like.

@@

# Supplement
## Supplemental figure 1

@@

## Supplemental figure 2

*Supplemental figure 2: Fraction of particulate carbon represented by known metabolites. Boxplots set behind points are colored according to sea level anomaly (SLA) with red corresponding to a positive SLA (anticyclone) and blue corresponding to negative SLA (cyclone).*

>>>>

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment [@Karl1999; @Karl2017; @Karl2021]. A major source of this variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016]. Prior research has established the importance of these mesoscale features in introducing nutrients into the euphotic zone and altering the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies, and @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation.

@@

*Figure 1: Plots of the metabolome across the eddy dipole, broken down by depth. The top row of plots are non-metric multidimensional scaling (NMDS) plots, in which points correspond to individual samples and have been colored by their corrected sea level height anomaly. NMDS stress values (s) have been reported in the bottom left corner, while PERMANOVA R$^2$ and p-values are reported in the top left. SLA trends are visible in the DCM and deep samples, with dark blue circles consistently discriminating from the dark red circles along the first multidimensional axis. The bottom row of plots show the direction and magnitude of this effect by plotting the grand median of the z-score normalized metabolite peak areas in the stations taken across the eddy dipole with the upper whisker extent, median bar, and lower whisker extent corresponding to the values measured in three biological triplicates.*

@@

```{r figure stacked barplot MT eddy transect w PC and SLA}
avg_targ_concs <- targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  group_by(compound_name) %>%
  mutate(mean_cmpd_nm=mean(nM)) %>%
  ungroup()
arrow_df <- data.frame(
  cyc_type=c("Anticyclone\ncenter", "Cyclone\ncenter"),
  x=c("6", "12"), xend=c("6", "12"),
  y=c(5, 4), yend=c(9.5, 6.5),
  color=c(scales::muted("red"), scales::muted("blue")),
  depth=factor("Deep", levels=c("Surface", "DCM", "Deep"), 
               labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)")),
  lab_x=c("5", "10")
)
barplot_targ_gp <- avg_targ_concs %>%
  group_by(compound_name, untripl) %>%
  summarise(nM=median(nM), mean_cmpd_nm=unique(mean_cmpd_nm)) %>%
  ungroup() %>%
  arrange(desc(mean_cmpd_nm)) %>%
  mutate(compound_name=ifelse(compound_name=="Trimethylamine N-oxide", "TMAO", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Hydroxyisoleucine", "HO-Ile", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Glycine betaine", "GBT", compound_name)) %>%
  mutate(compound_name=str_remove(compound_name, "^L-")) %>%
  mutate(compound_name=ifelse(
    compound_name%in%head(unique(compound_name), 9), 
    compound_name, "Other")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  left_join(filled_file_metadata %>% distinct(untripl, depth, station)) %>%
  group_by(depth, station, compound_name) %>%
  summarise(nM=sum(nM), .groups = "keep") %>%
  summarise(mean_nM=mean(nM)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=station, y=mean_nM, fill=compound_name), width = 0.9) +
  scale_fill_manual(values = c(splividis(9), "grey50")) +
  scale_y_continuous(expand = c(0.005, 0.1)) +
  labs(x=NULL, y="Median nM\nmetabolite concentration", fill=NULL) +
  facet_grid(~depth, scales="free", space = "free") +
  guides(fill=guide_legend(nrow=2, byrow = TRUE)) +
  ggnewscale::new_scale_fill() +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend, color=color), data=arrow_df,
               arrow=arrow(ends = "first", length = unit(0.1, "npc"), type = "closed"),
               linewidth=0, linejoin = "bevel") +
  geom_segment(aes(x=x, y=y+0.5, xend=xend, yend=yend, color=color), 
               data=arrow_df, linewidth=2) +
  geom_text(aes(x=lab_x, y=yend+0.5, color=color, label=cyc_type), 
            data=arrow_df, vjust=0, hjust=0, lineheight=0.8) +
  scale_x_discrete(breaks=as.character(seq(4, 14, 2))) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), guide = "none"
  ) +
  scale_color_identity() +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_blank(), 
        legend.margin = margin(t = 10,0,0,0), 
        legend.box.margin = margin(-10,-10,-10,-10),
        plot.margin = unit(numeric(4), "cm"),
        strip.text = element_text(face = "bold"))

frac_pc_in_metab_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity)) %>%
  mutate(n_C=as.numeric(str_extract(formula, "(?<=C)\\d+"))) %>%
  mutate(nM_C=n_C*nM) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  filter(samp_type=="Smp") %>%
  group_by(filename) %>%
  summarize(total_nM=sum(nM_C), PC_um=unique(PC_um)) %>%
  mutate(frac_in_metab=total_nM/1000/PC_um) %>%
  left_join(filled_file_metadata, by = join_by(filename, PC_um)) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))
pc_frac_metab_gp <- frac_pc_in_metab_df %>%
  ggplot(aes(x=station, y=frac_in_metab, color=sla_corr)) +
  geom_hline(yintercept = 0) +
  geom_boxplot(width = 0.9, linewidth=0.8) +
  scale_x_discrete(breaks=as.character(seq(4, 14, 2))) +
  scale_y_continuous(labels = scales::label_percent()) +
  facet_grid(~depth, scales="free") +
  scale_color_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill")
  ) +
  guides(color=guide_colorbar(direction = "horizontal", title.position = "top", 
                              title.hjust = 0.5, barheight = 0.2, reverse = TRUE)) +
  labs(color="Corr. SLA (cm)", x="Station",
       y="Fraction of PC represented\nby known metabolites") +
  theme_bw() +
  theme(legend.position = c(0.84, 0.75), legend.background = element_rect(color="grey90"),
        strip.background = element_blank(), strip.text = element_blank(),
        plot.margin = margin(0,1,0,0))

targ_gp_w_sla_frac <- egg::ggarrange(barplot_targ_gp, pc_frac_metab_gp, draw = FALSE)

no_outlier_frac_pc_in_metab_df <- frac_pc_in_metab_df %>%
  filter(!(station=="11" & tripl=="A" & depth=="Surface")) %>%
  filter(!(station=="6" & tripl=="B" & depth=="Surface")) %>%
  filter(!(station=="7" & tripl=="C" & depth=="Surface")) %>%
  filter(!(station=="11" & tripl=="B" & depth=="DCM"))
pc_totalnm_plot <- no_outlier_frac_pc_in_metab_df %>%
  ggplot() +
  aes(x=PC_um, y=total_nM) +
  geom_smooth(method="lm", color="black") +
  geom_point(aes(fill=sla_corr), pch=21, color="black") +
  facet_wrap(~depth, ncol=1, scales="free") +
  labs(x="Particulate\ncarbon (uM)", y="Total nM carbon in known metabolites") +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), guide = "none"
  ) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))

targ_gp_w_sla_pc <- cowplot::plot_grid(
  targ_gp_w_sla_frac, pc_totalnm_plot, rel_widths = c(3, 1), labels = "AUTO"
) +
  theme(plot.background = element_rect(fill="white", color=NA))

plot(targ_gp_w_sla_pc)

ggsave("targ_gp_w_sla_pc.png", plot = targ_gp_w_sla_pc, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300)
```

![](figures/targ_gp_w_sla_pc.png)

*Figure 2: A) Upper row of stacked barplots shows known compounds measured across the eddy dipole and separated by depth. Bar height corresponds to median triplicate concentration for each metabolite, with the top 9 shown and the 44 other identified metabolites summed in grey. TMAO = trimethylamine N-oxide, HO-Ile = hydroxyisoleucine, GBT = glycine betaine, DCM = deep chlorophyll maximum. Lower row of boxplots shows the fraction of total particulate carbon (PC) in the known metabolites across the eddy dipole, with the upper whisker extent, median bar, and lower whisker extent corresponding to the values measured in three biological triplicates. Colors correspond to corrected sea level anomaly (Corr. SLA) in centimeters, with dark red indicating anticyclonic (positive SLA) and dark blue indicating cyclonic (negative SLA) eddy state. B) Type I linear regressions of particulate carbon against total nM carbon in known metabolites. Points have been colored according to the SLA scheme in A) and have been placed on independent axes. Outlier points have been removed from the surface at stations 6, 10, and 11 as well as a DCM sample from station 12. For the full regression, see Supplemental Figure XX.*

```{r avg targ conc values and numbers for text, eval=FALSE}
avg_targ_concs %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  group_by(depth, compound_name) %>%
  summarise(nM=mean(nM)) %>%
  arrange(compound_name) %>%
  pivot_wider(names_from = depth, values_from = nM) %>%
  write_csv("supplement/avg_targ_conc_by_depth.csv")
avg_targ_concs %>%
  mutate(depth=factor(depth, levels=rev(c("Surface", "DCM", "Deep")))) %>%
  group_by(compound_name, depth) %>%
  summarise(nM=mean(nM)) %>%
  mutate(nM=log10(nM)) %>%
  ggplot(aes(x=nM, y=depth, group=compound_name)) +
  geom_path()
avg_targ_concs %>%
  mutate(depth=factor(depth, levels=rev(c("Surface", "DCM", "Deep")))) %>%
  group_by(compound_name, depth) %>%
  summarise(nM=mean(nM)) %>%
  mutate(nM=rank(nM)) %>%
  ungroup() %>%
  pivot_wider(names_from = depth, values_from = nM) %>%
  mutate(depth_class=paste(Surface, DCM, Deep)) %>%
  count(depth_class)
```

We found that the overall concentration of known molecules was highest at the DCM of the cyclone, in line with our expectations and prior results. The most abundant quantified molecule overall was trimethylamine N-oxide (TMAO), with an average concentration of 1.38 nM and a distinct DCM maximum (Figure 2). This pattern of a subsurface maximum at the DCM was mirrored by a majority of the known molecules (37/53). Other abundant molecules decreased monotonically with depth, with highest values at the surface (Figure 2).

```{r PC_vs_totnM regression values, eval=FALSE}
no_outlier_frac_pc_in_metab_df %>%
  ggplot(aes(x=PC_um, y=total_nM)) + 
  geom_smooth(method="lm", color="grey40") +
  geom_point(aes(color=depth)) + 
  geom_smooth(aes(color=depth), method="lm") +
  theme_bw()

# Overall trend between PC and total nM metabolites (all depths)
broom::tidy(lm(total_nM~PC_um, data=no_outlier_frac_pc_in_metab_df))[2,]
with(no_outlier_frac_pc_in_metab_df, cor(PC_um, total_nM))
lme4::lmer(total_nM~PC_um + (1|depth), data=no_outlier_frac_pc_in_metab_df)

# By-depth trend between PC and total nM metabolites
no_outlier_frac_pc_in_metab_df %>%
  group_by(depth) %>%
  summarise(overall=broom::tidy(lm(total_nM~PC_um))[2,], 
            cor_vals=cor(PC_um, total_nM))
```

As expected, bulk particulate carbon (PC) measurements showed a strong trend with summed metabolite concentrations ($\beta = 23.7$, Pearson's $r$ = .876, p-value < 0.001). This overall trend was largely driven by both higher values in the surface than at depth but when analyzed at each depth individually the general correspondence held. The DCM had the strongest trend ($\beta$ = 22.1, $r= .689$, p-value < 0.001) while Surface ($\beta$ = 16.0, $r = .375$, p-value = 0.04) and Deep ($\beta$ = 23.9, $r = 0.503$, p-value = 0.003) values were weaker but still compelling relationships. However, this association required the removal of three samples in the surface and one in the DCM with anomalously high metabolite loads. The full regression is reported in Supplemental Figure XX.

```{r PC vs total_nM ANOVA by depth, eval=FALSE}
# More carbon in metabolites at DCM when outliers removed?
no_outlier_frac_pc_in_metab_df %>%
  mutate(pred_total_nM=predict(lm(formula=total_nM~PC_um, data=.), newdata=.)) %>%
  select(filename, depth, station, sla_class, sla_corr, total_nM, PC_um, pred_total_nM) %>%
  mutate(total_nM_resid=total_nM-pred_total_nM) %>%
  aov(formula=total_nM_resid~depth) %>% TukeyHSD() %>% broom::tidy()
  # ggplot() + geom_boxplot(aes(x=depth, y=total_nM_resid)) + aes(color=sla_class)

# More carbon in metabolites at DCM when outliers handled with rank testing?
frac_pc_in_metab_df %>%
  mutate(pred_total_nM=predict(lm(formula=total_nM~PC_um, data=.), newdata=.)) %>%
  select(filename, depth, total_nM, PC_um, pred_total_nM) %>%
  mutate(total_nM_resid=rank(total_nM-pred_total_nM)) %>%
  kruskal.test(formula=total_nM_resid~depth) %>% broom::tidy()
  # ggplot() + geom_boxplot(aes(x=depth, y=total_nM_resid))

# More carbon in metabolites at DCM when measured by % in metabs too
no_outlier_frac_pc_in_metab_df %>%
  aov(formula=frac_in_metab~depth) %>% TukeyHSD() %>% broom::tidy()
  # ggplot() + geom_boxplot(aes(x=depth, y=frac_in_metab))
no_outlier_frac_pc_in_metab_df %>%
  group_by(depth) %>%
  summarise(median(frac_in_metab))
frac_pc_in_metab_df %>%
  kruskal.test(formula=frac_in_metab~depth, data=.) %>% broom::tidy()
  # ggplot() + geom_boxplot(aes(x=depth, y=rank(frac_in_metab)))
```

Known metabolites contributed between 1.5% and 5% of the carbon to the total PC load with higher values at the DCM than the surface or 175m samples (Figure 2, Supplemental Figure XX). DCM samples contained approximately 25% (0.5 percentage points) more carbon in the known metabolite pool (One-way ANOVA with outliers removed plus Tukey post-hoc, p<0.001, n=30-33) but there was no clear trend with SLA.

```{r comparison to Angies PC values way higher than mine, eval=FALSE}
angie_pcdata <- "../metadata/raw_data/Karl_HL2_WC_PCPN_Final_Data.xlsx" %>%
  openxlsx::read.xlsx() %>%
  slice(-1) %>%
  select(abs_depth=`Depth.(m)`, PC_um=`C.value.(µmol/L)`) %>%
  # filter(depth<30) %>%
  mutate(cruise="HOE-Legacy")

my_pcdata <- filled_file_metadata %>%
  # filter(depth=="Surface") %>%
  distinct(abs_depth, PC_um) %>%
  mutate(cruise="MESOSCOPE")

rbind(angie_pcdata, my_pcdata) %>%
  ggplot() +
  geom_point(aes(x=PC_um, y=-abs_depth, color=cruise))
```

@@

@@

# Supplement
## Supp fig: PCA axes 1-3 vs SLA by depth

@@

## Supp fig: Stacked barplot of top metabs by triplicate
